### 2025-07-12 multiple-regression.qmd
20250712 - heteroscedasticity types

```{r}
library("petersenlab")
library("rms")
library("car")
library("caret")
library("lme4")
library("performance")
library("lavaan")
library("mice")
library("miceadds")
library("interactions")
library("brms")
library("robustbase")
library("ordinal")
library("MASS")
library("broom")
library("effectsize")
library("tidyverse")
library("knitr")
```

@@

```{r}
#| layout: [[100], [50, 50]]
#| label: fig-homscedasticityHeteroscedasticity
#| fig-cap: "Example of Homoscedasticity and Heteroscedasticity in Residual Plots."
#| fig-alt: "Example of Homoscedasticity and Heteroscedasticity in Residual Plots."
#| fig-subcap:
#|   - "Homoscedasticity"
#|   - "Fan-Shaped Heteroscedasticity"
#|   - "Bow-Tie-Shaped Heteroscedasticity"
#| code-fold: true

# 1. Homoscedasticity
sampleSize <- 1000
x1 <- runif(sampleSize, 0, 10)
y1 <- 3 + 2 * x1 + rnorm(sampleSize, mean = 0, sd = 2)
fit1 <- lm(y1 ~ x1)
res1 <- resid(fit1)
fitted1 <- fitted(fit1)

@@

plot(
  fitted1,
  res1,
  main = "Homoscedasticity",
  xlab = "Fitted value",
  ylab = "Residual")

@@

>>>>

```{r}
library("petersenlab")
library("rms")
library("car")
library("caret")
library("lme4")
library("performance")
library("lavaan")
library("mice")
library("miceadds")
library("interactions")
library("brms")
library("robustbase")
library("ordinal")
library("MASS")
library("broom")
library("effectsize")
library("tidymodels")
library("tidyverse")
library("knitr")
```

@@

```{r}
#| layout: [[100], [50, 50], [50, 50]]
#| label: fig-homscedasticityHeteroscedasticity
#| fig-cap: "Example of Homoscedasticity and Heteroscedasticity in Residual Plots."
#| fig-alt: "Example of Homoscedasticity and Heteroscedasticity in Residual Plots."
#| fig-subcap:
#|   - "Homoscedasticity"
#|   - "Fan-Shaped Heteroscedasticity"
#|   - "Bow-Tie-Shaped Heteroscedasticity"
#|   - "Diamond-Shaped Heteroscedasticity"
#|   - "Triangle-Shaped Heteroscedasticity"
#| code-fold: true

set.seed(52242)
sampleSize <- 1000

# 1. Homoscedasticity
x1 <- runif(sampleSize, 0, 10)
y1 <- 3 + 2 * x1 + rnorm(sampleSize, mean = 0, sd = 2)
fit1 <- lm(y1 ~ x1)
res1 <- resid(fit1)
fitted1 <- fitted(fit1)

@@

# 4. Diamond-shaped heteroscedasticity
x4 <- runif(sampleSize, 0, 10)
sd4 <- -abs(x4 - 5) + 5.5  # inverted bow-tie
y4 <- 3 + 2 * x4 + rnorm(sampleSize, mean = 0, sd = sd4)
fit4 <- lm(y4 ~ x4)
res4 <- resid(fit4)
fitted4 <- fitted(fit4)

#5. Triangle-shaped heteroscedasticity
height <- 10

sample_triangle <- function(n, v1, v2, v3) {
  u <- runif(n)
  v <- runif(n)
  is_flip <- u + v > 1
  u[is_flip] <- 1 - u[is_flip]
  v[is_flip] <- 1 - v[is_flip]

  x <- (1 - u - v) * v1[1] + u * v2[1] + v * v3[1]
  y <- (1 - u - v) * v1[2] + u * v2[2] + v * v3[2]

  cbind(x, y)
}

# Triangle vertices
v1 <- c(0, height)  # top left
v2 <- c(10, height) # top right
v3 <- c(5, 0)       # bottom center

# Sample points inside triangle
tri_points <- sample_triangle(
  sampleSize,
  v1,
  v2,
  v3)

# Extract x and residuals
x5 <- tri_points[, 1]
residuals5 <- tri_points[, 2]

# Generate outcome
y5 <- 3 + 2 * x5 + residuals5

# Fit model
fit5 <- lm(y5 ~ x5)
res5 <- resid(fit5)
fitted5 <- fitted(fit5)

# Plots
plot(
  fitted1,
  res1,
  main = "Homoscedasticity",
  xlab = "Fitted value",
  ylab = "Residual")

@@

plot(
  fitted4,
  res4,
  main = "Diamond-Shaped Heteroscedasticity",
  xlab = "Fitted value",
  ylab = "Residual")

plot(
  fitted5,
  res5,
  main = "Triangle-Shaped Heteroscedasticity",
  xlab = "Fitted value",
  ylab = "Residual")
